
---
title: "Sparse matrix examples"
author: "Steve Simon"
date: "Sunday, April 17, 2016"
output: html_document
---

```{r start-clean}
# backup, just in case
save.image("backup.RData")
rm(list=ls())
library("Matrix")
library("knitr")
knit_hooks$set(timer = function(before, options, envir) {
  if (before) {
    current_time <<- Sys.time()
  }
  else {
    elapsed_time <- Sys.time() - current_time
    elapsed_message <- paste(round(elapsed_time, 1), attr(elapsed_time, "units"))
    message(elapsed_message)
    m <- paste("Chunk", options$label, "used", elapsed_message, "\n")
    if (exists("timing_log")) {
      timing_log <<- paste(timing_log, m, sep="\n")
    } else {
      timing_log <<- m
    }
    return(timing_log)
  }
})
timer <- function(initialize=FALSE) {
  if (initialize) {
    tm <<- Sys.time()
    cat("\n\nProgram started:", as.character(tm), "\n", sep="")
    return(invisible(tm))
  }
  tm <<- c(tm, Sys.time())
  n <- length(tm)
  et <- round(tm[n] - tm[n-1])
  es <- paste(et, attr(et, "units"))
  cat("\n\nProgram time:", es)
  return(invisible(tm))
}
tst1 <- timer(TRUE)
tst2 <- timer()
tst3 <- timer() 
```

Set up a matrix where all the entries are zero, except for
every power of 2.

```{r create a simple matrix, message=TRUE, timer=TRUE}
message("Test 1")
nt <- 7
nc <- 3
v <- rep(0,2^nt)
v[2^(0:nt)] <- 1
m <- matrix(v, ncol=2^nc)
print(m)
str(m)
```

This is a regular matrix with `r 2^nt` entries arranged in
`r 2^(nt-nc)` rows and `r 2^nc` columns. Exactly `r nt+1`
values are non-zero. 

A sparse matrix is a matrix that only stores the non-zero values
and their locations. If the matrix is very large and most of
the values are zero, then you can represent even a very very large
matrix as a sparse matrix without exceeding the memory limits of R.

To create a sparse matrix, you need a special library. The one 
you see here, Matrix, is a popular one that is supported
by (among other things) the glmnet package.

To create a sparse matrix, use the Matrix function instead of
the matrix function.

```{r create sparse matrix, message=FALSE, timer=TRUE}
message("Test 2")
s <- Matrix(v, ncol=2^nc)
print(s)
str(s)
```

The sparse matrix is stored as an S4 object. The pieces of the S4
object are a bit tricky to interpret. The values stored in  i: `r s@i`
tell you which rows the non-zero values are stored in. The trick here
is that Matrix adopts the convention of C++ and many other languages
that the first row (and the first column) of a matrix start at zero.
In R, the first row (and the first column) of a matrix starts at one.

There are some good reasons why you might prefer the C++ numbering convention
and some good reasons why you might prefer the R numbering convention. If it
helps, you can list the rows in R format by adding one: `r s@i+1`.

You can use the dimnames funtion to label the rows and columns of s in the C++
numbering convention.

```{r relabel sparse matrix, timer=TRUE}
dimnames(s) <- list(0:(2^(nt-nc)-1), 0:(2^nc-1))
s
```

Now knowing which row is only half the battle. You also need to know which column.
The columns are designated by a pointer stored in p: `r s@p`. Pointers are 
commonly used in C++, but are an alien concept to most R programmers (or maybe
just this R programmer).

The first two pointers: `r s@p[1:2]` tell you that the rows listed in i, starting
at the `r s@p[1]` location and going up to but not including the `r s@p[2]` position
are rows in the first column. Remember that Matrix, like C++, starts counting at 0,
not 1. So the first column has entries in the following rows: `r s@i[(s@p[1]+1):s@p[2]]`.

The second and third pointers: `r s@p[2:3]` tell you that the fifth row position: `r s@i[s@p[2]+1]`
is in the second column. Since the third and the third pointers: `r s@p[3:4]` are the same, there
are no non-zero entries in the third row.

Are you confused by all this? I know I am. The help file, though, has a nice hint. Calculate the
difference between successive pointer values: `r diff(s@p)`. This will tell you that the first 
column has `r diff(s@p)[1]` entries, the second column has `r diff(s@p)[2]` entry, the third
column has `r diff(s@p)[3]` entries and so forth.

You can use the rep function to compute the column indices that go with the row indices.

```{r calculate column indices, timer=TRUE}
j <- rep(0:(2^nc-1), diff(s@p))
print(data.frame(i=s@i, j))
```

You can create a sparse matrix by specifying the non-zero rows and columns.

```{r specify rows and columns, timer=TRUE}
row.indices <- s@i+1
col.indices <- j+1
sparseMatrix(i=row.indices, j=col.indices, dims=c(2^(nt-nc), 2^nc), x=1)
```

The values stored in x: `r s@x` are all ones because in this example, all the non-zero entries are one.
That's not always the case, which is why the sparse matrix stores the individual non-zero values. 
This is a pattern matrix, a sparse matrix where all the values are either zero or one (mostly zeros,
of course, or the matrix isn't sparse). You might be able to improve slightly on 

The regular matrix takes up `r object.size(m)` bytes. The sparse matrix takes up `r object.size(s)` bytes. That seems odd,
but the sparse matrix format only makes sense when the matrices get really really large.

```{r examine sparse matrix, timer=TRUE}
for (i in 1:5) {
  nt <- 7+3*i
  nc <- 3+i
  v <- rep(0,2^nt)
  v[2^(0:nt)] <- 1
  m <- matrix(v, ncol=2^nc)
  s <- Matrix(v, ncol=2^nc)
  cat("\n\nA regular", 2^(nt-nc), "by", 2^nc, "matrix requires", object.size(m), "bytes.")
  cat("\nA sparse", 2^(nt-nc), "by", 2^nc, "matrix requires", object.size(s), "bytes.")
}
```

Save everything for possible later use.

```{r save-everything, timer=TRUE}
save.image("sparse.RData")
```
